---
title: "Conductivity Investigation Winter 2022"
author: "Katheryn Rein"
date: "2024-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE, message = FALSE, fig.width=5, fig.height=5)
```

#### This is a R Markdown HTML document to supplement DWR-1-INV-008 Conductivity Investigation Winter 2022 which explores cleaned up data compiled from samples collected between March 2018 and February 2023 that had both field and laboratory derived conductivity reported in the DWR Water Data Library (WDL). Relative percent differences (RPDs) betweene these field and laboratory reported values are evaluated here. explore relative percent differences (RPD) between DWR field and Bryte Lab specific conductance (SC) measurements. Samples collected between March 2018 and February 2023 are evaluated here. Data were exported from the WDL database. See xyz script for data cleaning and reorganization.


```{r, echo = FALSE}
library(writexl)
library(tidyverse)
library(data.table)
library(readxl)
library(janitor)
library(ggpubr)
library(lubridate)
library(flextable)
library(beanplot)
library(FSA)
library(effects)
library(gplots)
library(car)
```

##### Format data to evaluate differences between time periods of interest.

```{r}

df <- read_excel("~/Projects/Field Lab SC/RPD calculation files from R/RPD data from full data set_v2.xlsx")

df$collection_date2 <- as.POSIXct(df$collection_date2, format="%m/%d/%Y %H:%M")

df = mutate(df, time_period_focus = case_when(collection_date2 >= as.Date("2022-11-01")~"1", 
          collection_date2 < as.Date("2022-11-01") &
          collection_date2 >= as.Date("2022-07-01")~"2")) 
                                         

df$time_period_focus <- as.numeric(df$time_period_focus)


#remove data for data owners 3467 and 6200 since neither have data in both time period 1 and time period 2.This removes 25 samples with 4189 remaining.

df_1 = filter(df, data_owner_id=="6301"| data_owner_id=="1860"| data_owner_id=="2040"| data_owner_id=="0310"|
                data_owner_id=="4804"|data_owner_id=="6301" |data_owner_id=="7200")

#remove data not in time period 1 or 2, get a total of 918 samples
df_1_focus = filter(df_1, time_period_focus=="1" | time_period_focus=="2" )

#summary of time periods 1 & 2

time_dates <- df_1_focus %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_date = min(as.Date(collection_date2)),
            max_date = max(as.Date(collection_date2)))

tableA <- flextable(time_dates)

tableA <- width(tableA, width = 1.5)

tableA <- theme_vanilla(tableA)
tableA <- set_caption(tableA, caption = "Table A. Sample counts and date ranges for time periods of focus")

tableA
```

#### The RPD across the periods 1 & 2 (Fig. 1, Fig. 2) is not normally distributed (Fig. 3).The Shapiro-Wilk test did not find normality in the data or two transformations of it (p \< 0.05). Data were filtered above and below 10,000 µS/cm and the normality of each dataset was also not normal.
#??Higher RPDs are more commonly observed at specific conductance (SC) measurements \< 10,000 µS/cm (Fig. 4, Fig. 5). ??

#### High RPDs are observed at SC within drinking water standards (200 - 800 µS/cm) and through brackish conditions (\<10,000 µS/cm). The median SC of the dataset (\~500 µS/cm) is less than the mean value (\~2950 µS/cm) identifying a skewness in the the dataset that is the likely cause that the data is not normal.???

#### Conductivity sensors used in the field are not commonly calibrated below 1000 microseimens per centimeter (µS/cm) per the manufacturers recommendations. Current department guidance includes field programs checking the conductivity sensor to ensure the sensor reads \< 5 µS/cm in air.

```{r}

ggplot(df_1_focus, aes(y = rpd, x = collection_date2)) +
  geom_point() +
  ggtitle("Figure 1. The RPD of Field and Lab SC Results across time periods 1 & 2 ") +
  ylab("RPD (%)") + 
  xlab("Collection date")

 
hist(df_1_focus$rpd, breaks=100, main = "Figure 2. Conductivity RPD distribution") 

qqnorm(df_1_focus$rpd, main = "Figure 3. Normal Q-Q Plot")

qqline(df_1_focus$rpd, main = "Figure 3. Normal Q-Q Plot") #this doesn't work is this needed?

summary(df_1_focus$field_result)

summary(df_1_focus$lab_result)

summary(df_1_focus$rpd)
  
shapiro.test(df_1_focus$rpd) #full focus data set is not normal, p<2.2e-16

ggplot(df_1_focus, aes(y = rpd, x = field_result)) +
  geom_point() +
  ylab("RPD (%)") + 
  xlab("Field SC (uS/cm)")+
  ggtitle("Figure 4. RPD of Field SC") +
  facet_grid(~time_period_focus) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Filter data to below 10,000 uS/cm

df_1_focus_below10K <- df_1_focus %>% 
  filter(field_result < 10000 & lab_result < 10000) #838/918 samples have field and lab result <10,000 uS/cm

shapiro.test(df_1_focus_below10K$rpd) #data below 10K uS/cm are not normal across time periods 1 & 2, p<2.2e-16

hist(df_1_focus_below10K$rpd, breaks = 100, main = "Figure 5. RPD below 10,000")

# Filter data to above 10,000 uS/cm

df_1_focus_above10K <- df_1_focus %>% 
  filter(field_result >= 10000 | lab_result >=10000) #80/918 samples have field and/or lab result > or = 10,000 uS/cm

hist(df_1_focus_above10K$rpd, breaks = 100, main = "Figure 6. RPD above 10,000")

shapiro.test(df_1_focus_above10K$rpd) #data at or above 10K uS/cm are not normal, p=3.399e-07

#Plot values < 10000 uS/cm
ggplot(df_1_focus, aes(y = rpd, x = field_result)) +
  geom_point() +
  ylab("RPD (%)") + 
  xlab("SC (uS/cm)")+
  xlim(0,10000)+
  ggtitle("Figure 7. RPD of Field SC (< 10000 uS/cm)") +
  facet_grid(~time_period_focus) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#Plot values =< 10000 uS/cm
ggplot(df_1_focus, aes(y = rpd, x = field_result)) +
  geom_point() +
  ylab("RPD (%)") +
  ylim(-5,5)+
  xlab("SC (uS/cm)")+
  xlim(10000,50000)+
  ggtitle("Figure 8. RPD of Field SC (>= 10000 uS/cm)") +
  facet_grid(~time_period_focus) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#transformations run that did not yield a normally-distributed dataset.
#df <- df %>% 
  #mutate(rpd_sqr = rpd^2,
         #log_abs_rpd = log(abs(rpd)+0.1))

#shapiro.test(df$rpd_sqr)

#shapiro.test(df$log_abs_rpd)
```

#### Summary tables of sample counts and RPDS across time periods 1 & 2 and filtered for above and below 10000 uS/cm

```{r}


time2 <- df_1_focus %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))


table1 <- flextable(time2)

table1 <- theme_vanilla(table1)
table1 <- set_caption(table1, caption = "Table 1. SC Measurements across time periods 1 & 2")

table1

#data < 10000 uS/cm between time periods 1 & 2

time2_below10K <- df_1_focus_below10K %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))

table2 <- flextable(time2_below10K)

table2 <- theme_vanilla(table2)
table2 <- set_caption(table2, caption = "Table 2. SC Measurements < 10,000 uS/cm across time periods 1 & 2")

table2

#data >= 10000 uS/cm across time periods 1 & 2

time2_above10K <- df_1_focus_above10K %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))

table3 <- flextable(time2_above10K)

table3 <- theme_vanilla(table3)
table3 <- set_caption(table3, caption = "Table 3. SC Measurements > 10,000 uS/cm across time periods 1 & 2")

table3

time_grouped <- df_1_focus %>%
  group_by(data_owner_id, time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_SC = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))

table5 <- flextable(time_grouped)

table5 <- theme_vanilla(table5)
table5 <- set_caption(table5, caption = "Table 5. Conductivity Summary by Group and Time Period")

table5

```


```
##### A Kruskal-Wallis rank sum test was used to assess statistical significance of RPDs between the two time periods. This is a nonparametric hypothesis test, similar to a one-way ANOVA) identified significant differences between the two time periods only when the field and/or lab conductivity value were >10,000 uS/cm. 

```{r}
ggboxplot(df_1_focus, x = "time_period_focus", y = "rpd", 
          color = "time_period_focus",
          ylab = "RPD (%)", xlab = "time period") +
  theme(text=element_text(size=20)) +
  ggtitle("Figure 7. RPD across timeperiods")

beanplot(rpd ~ time_period_focus, data = df_1_focus, col = c("#CAB2D6", "#33A02C", "#B2DF8A"), border = 'blue', #border = "#CAB2D6', 
         axes = TRUE,
         xlab = "timeperiod", ylab = "RPD (%)")

#test for statistical significance in RPD between time periods 1 & 2
kruskal.test(rpd ~ time_period_focus, data = df_1_focus) #not significant, p=0.2868

#values < 10000
kruskal.test(rpd ~ time_period_focus, data = df_1_focus_below10K) #not significant, p=0.5718

#values >= 10000
kruskal.test(rpd ~ time_period_focus, data = df_1_focus_above10K) #significant, p=4.81e-5

```

### Look for statistical difference within each field program between time periods 1 & 2

```{r}
#table by data owner

dataowner <- df_1_focus %>%
  group_by(data_owner_id,
           time_period_focus) %>%
  summarize(n = n(),
            min_SC = round(min(field_result),0),
            max_SC = round(max(field_result),0))

dataowner1 <- dataowner %>% 
  pivot_wider(names_from = time_period_focus,
              values_from = c(n, min_SC, max_SC))

table4 <- flextable(dataowner1)

table4 <- theme_vanilla(table1)
table4 <- set_caption(table1, caption = "Table 4. Sample counts across time periods 1 & 2 by data owner ")

table4

#"0310" "1860" "2040" "3467" (no samples in both time periods) "4804" "6301" "7200" "6200" (no samples in both time periods)

#data owner 0310 (all subsets not normal)-----------------
#timeperiods 1 and 2 not significant
#<10K tp 1 and 2 not significant
#>10K tp 1 and 2 is significant

df_0310 <- df_1_focus %>% 
  filter(data_owner_id == "0310")

summary(df_0310)

df_0310_below10K <- df_1_focus_below10K %>% 
   filter(data_owner_id == "0310")

df_0310_above10K <- df_1_focus_above10K %>% 
   filter(data_owner_id == "0310")

shapiro.test(df_0310$rpd) #not normal, p<2.2e-16
shapiro.test(df_0310_below10K$rpd) #not normal, p<2.2e-16
shapiro.test(df_0310_above10K$rpd) #not normal, p=3.691e-7

kruskal.test(rpd ~ time_period_focus , data = df_0310) #not significant, p=0.5305

kruskal.test(rpd ~ time_period_focus , data = df_0310_below10K) #not significant, p=0.7814

kruskal.test(rpd ~ time_period_focus , data = df_0310_above10K) #significant, p=5.431e-05

time_0310_above10K <- df_0310_above10K %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))
            
time_0310 <- df_0310 %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))


#data owner 1860 - not normal, no values >10000 collected in time periods 1 or 2---------------------
#timeperiods 1 and 2 significant
#<10K tp 1 and 2 significant
#>10K tp 1 and 2 N/A

df_1860 <- df_1_focus %>% 
  filter(data_owner_id == "1860")

summary(df_1860)

shapiro.test(df_1860$rpd) #not normal, p=4.226e-14

kruskal.test(rpd ~ time_period_focus , data = df_1860) #significant, p=0.01245

time_1860 <- df_1860 %>%
  group_by(time_period_focus) %>%
  summarize(n = n(),
            min_field_SC = round(min(field_result), 0),
            max_field_SC = round(max(field_result),0),
            min_lab_SC = round(min(lab_result),0),
            max_lab_sc = round(max(lab_result),0),
            min_rpd = round(min(rpd),2),
            max_rpd = round(max(rpd),2),
            mean_rpd = round(mean(rpd), 2),
            median_rpd = round(median(rpd), 2))


#data owner 2040 - not normal, no values > 10000 collected in time periods 1 or 2-----------------------
#timeperiods 1 and 2 significant
#<10K tp 1 and 2 significant
#>10K tp 1 and 2 N/A

df_2040 <- df_1_focus %>% 
  filter(data_owner_id == "2040")

summary(df_2040)

shapiro.test(df_2040$rpd) #not normal, p<2.2e-16

kruskal.test(rpd ~ time_period_focus , data = df_2040) #significant, p=9.448e-4


#data owner 4804 - not normal, no values > 10000 collected in time periods 1 or 2 --------------------------
#timeperiods 1 and 2 significant
#<10K tp 1 and 2 significant
#>10K tp 1 and 2 N/A

df_4804 <- df_1_focus %>% 
  filter(data_owner_id == "4804")

summary(df_4804)

shapiro.test(df_4804$rpd) #not normal, p<2.2e-16

kruskal.test(rpd ~ time_period_focus , data = df_4804) #significant, p=1.888e-12


#data owner 6301 - not normal, no values > 10000 collected in time periods 1 or 2 --------------------------
#timeperiods 1 and 2 significant
#<10K tp 1 and 2 significant
#>10K tp 1 and 2 N/A

df_6301 <- df_1_focus %>% 
  filter(data_owner_id == "6301")

summary(df_6301)

shapiro.test(df_6301$rpd) #not normal, p=3.493e-10

kruskal.test(rpd ~ time_period_focus , data = df_6301) #significant, p=0.009539

#data owner 7200 - not normal for full range & below 10K, normal >= 10K--------------------------
#timeperiods 1 and 2 significant
#<10K tp 1 and 2 significant
#>10K tp 1 and 2 not signficant but on the cusp with very small sample size

df_7200 <- df_1_focus %>% 
  filter(data_owner_id == "7200")

summary(df_7200)

df_7200_below10K <- df_1_focus_below10K %>% 
   filter(data_owner_id == "7200")

df_7200_above10K <- df_1_focus_above10K %>% 
   filter(data_owner_id == "7200")


shapiro.test(df_7200$rpd) #not normal, p=8.826e-08
shapiro.test(df_7200_below10K$rpd) #not normal, p=5.539e-07
shapiro.test(df_7200_above10K$rpd) #normal, p=0.4983

kruskal.test(rpd ~ time_period_focus , data = df_7200) #significant, p=0.01218

kruskal.test(rpd ~ time_period_focus , data = df_7200_below10K) #significant, p=0.006142

model <- aov(rpd ~ time_period_focus, data = df_7200_above10K) #not significant but barely, p=0.053, only 5 samples though
summary(model)
